Speeding up eQTL scans in the BXD population using GPU

### 2019-10-11 paper.tex
add genome scan including covariates

\begin{eqnarray*}
	y_i &=& X_j \beta_j+ \epsilon_i,
	\quad \epsilon_i \sim N(0,\sigma_i^2I),
\end{eqnarray*}
where ${X}_j$ is a matrix that may include the $y-$intercept with
or without covariate(s), and the $j$-th candidate genetic marker
($j=1,\ldots,p$).  ${\beta}_j$ is a vector of the $j$-th eQTL
effects, and ${\epsilon}_i$ is random error distributed as $N(0,\sigma_i^2I)$.  Suppose $RSS_{0i}$
is a residual sum of squares under the null hypothesis of no eQTL, and $RSS_{1ij}$ is the residual sum of squares under the
alternative of existing eQTL at the $i$-th trait and $j$-th marker.  Then the LOD
score for the corresponding pair can be written as:

\begin{eqnarray*}
	LOD_{ij} &=& \frac{n}{2} \log_{10} \left( \frac{RSS_{0i}}{RSS_{1ij}} \right)\\
	&=& \frac{n}{2} \log_{10} \left( 1{-}r_{ij}^2 \right),
\end{eqnarray*}
where $r_{ij}$ is the correlation between the $i$-th trait and
$j$-th marker. Note that this only works for one-df tests.  If $Y^{\ast}$ and $G^{\ast}$ are respectively standardized 
trait ($Y$) and genotype ($G$) matrices, then the
correlation matrix is simply $$R=Y ^{\ast '}G^{\ast}.$$
%Indicate which statistical analysis has been performed and describe the method and model applied. If many genes were examined simultaneously, or many phenotypes, a multiple comparison correction should be used to control the type I error rate, or a rationale for not applying a correction must be provided. The type of correction applied should be clearly stated. It should also be clear whether the p-values reported are raw, or after correction. Corrected p-values are often appropriate, but raw p-values should be available in the supporting materials so that others may perform their own corrections.

\subsection{Acceleration Techniques}
While it is true that many programs can achieve tens or even hundreds of speedup by utilizing GPU, the difference needs to be examined with more care. 
Sometimes, the reported CPU time is using single thread, and multithreaded CPU time may bring the performance gap of CPU and GPU closer than claimed. 
Also depending on the library chozen for CPU, speed might vary depending on whether the library is optimized. 
We believe for a fair comparison, both CPU and GPU functions should be optimized at maximum speed. 
The following section explains our efforts of optimization on CPU and GPU functions.

>>>>

\begin{eqnarray*}
	y_i &=& X_j \beta_j+ \epsilon_i,
	\quad \epsilon_i \sim N(0,\sigma_i^2I),
\end{eqnarray*}
where ${X}_j$ is a matrix that may include the $y$-intercept with
or without covariate(s), and the $j$-th candidate genetic marker
($j=1,\ldots,p$).  ${\beta}_j$ is a vector of the $j$-th eQTL
effects, and ${\epsilon}_i$ is random error distributed as $N(0,\sigma_i^2I)$.  Suppose $RSS_{0i}$
is a residual sum of squares under the null hypothesis of no eQTL, and $RSS_{1ij}$ is the residual sum of squares under the
alternative of existing eQTL at the $i$-th trait and $j$-th marker.  Then the LOD
score for the corresponding pair can be written as:

\begin{eqnarray*}
	LOD_{ij} &=& \frac{n}{2} \log_{10} \left( \frac{RSS_{0i}}{RSS_{1ij}} \right)\\
	&=& \frac{n}{2} \log_{10} \left( 1{-}r_{ij}^2 \right),
\end{eqnarray*}
where $r_{ij}$ is the correlation between the $i$-th trait and
$j$-th marker. Note that this only works for one-df tests.  If $Y^{\ast}$ and $G^{\ast}$ are respectively standardized 
trait ($Y$) and genotype ($G$) matrices, then the
correlation matrix is simply $$R=Y ^{\ast '}G^{\ast}.$$

Genome scan including covariates becomes slightly different for efficient implementation.  The idea is to project genetic markers 
and gene expressions onto the space generated by covariates and to compute the corresponding correlation matrix just as did for one-df tests.  In other words, 
let $Z$ be a matrix of covariates including $y$-intercept.  The orthogonal projector onto the covariate space is then $P_Z= Z(Z'Z)^{-1}Z'$.  
The markers ($G$) and gene expressions ($Y$) are now $Y_p =P_ZY$, $G_p= P_ZG$, respectively.  Standardizing them can compute the correlation matrix as shown above.

%Indicate which statistical analysis has been performed and describe the method and model applied. If many genes were examined simultaneously, or many phenotypes, a multiple comparison correction should be used to control the type I error rate, or a rationale for not applying a correction must be provided. The type of correction applied should be clearly stated. It should also be clear whether the p-values reported are raw, or after correction. Corrected p-values are often appropriate, but raw p-values should be available in the supporting materials so that others may perform their own corrections.

\subsection{Acceleration Techniques}
While it is true that many programs can achieve tens or even hundreds of speedup by utilizing GPU, the difference needs to be examined with more care. 
Sometimes, the reported CPU time is using single thread, and multithreaded CPU time may bring the performance gap of CPU and GPU closer than claimed. 
Also depending on the library chosen for CPU, speed might vary depending on whether the library is optimized. 
We believe for a fair comparison, both CPU and GPU functions should be optimized at maximum speed. 
The following section explains our efforts of optimization on CPU and GPU functions.
